\chapter{Scale of the solar system and local stellar environment}

\begin{quotation}
	\textit{Space is big. You just won't believe how vastly, hugely, mind-bogglingly big it is. I mean, you may think it's a long way down the road to the chemist's, but that's just peanuts to space.} \sourceatright{Douglas Adams, The Hitchhiker's Guide to the Galaxy}
\end{quotation}

Why can't we just see planets orbiting other stars with normal observational techniques like looking through larger and larger telescopes? Through this lab, we hope you gain a felt sense appreciation for the scale of star systems and the distance between the Sun and our closest stellar neighbor, and see why we need to use special techniques to detect exoplanets.

\section{Forming Groups}

If you are attending the lab session live and do not yet have a group, one way the TA could assist is to arrange "speed networking" among those who still need a group. This would involve the TA organizing Zoom Breakout Rooms, where each room is 2-3 students, and each group talks about how they work and what they are looking for in a group member. Then after 5 minutes or so, the Rooms are changed so people are with different people. This could help people get to know each other enough to form lab groups.

\begin{steps}
	\item Once you have a group, meet with each other and decide a) what tools you will use to communicate and collaborate, b) when you will meet, c) what you will do when you need to change an agreement, and d) what you will do when a member has a concern about how the group is functioning. \textbf{Record your agreements.}% This part counts as data collection and analysis, so it can be identical in each member's report.}
\end{steps}

\subsection{Team roles}

\begin{steps}
	\item \textbf{Decide on roles} for each group member.
\end{steps}

The available roles are:
\begin{itemize}
	\item Facilitator: ensures time and group focus are efficiently used
	\item Scribe: ensures work is recorded
	\item Technician: oversees apparatus assembly, usage
	\item Skeptic: ensures group is questioning itself
\end{itemize}

These roles can rotate each lab, and you will report at the end of the lab report on how it went for each role. Some members will be holding more than one role. For example, you could have the skeptic double with another role. Consider taking on a role you are less comfortable with, to gain experience and more comfort in that role.

Additionally, if you are finding the lab roles more restrictive than helpful, you can decide to co-hold some or all roles, or think of them more like functions that every team needs to carry out, and then reflecting on how the team executed each function.

\subsection{Add members to Canvas lab report assignment group}

\begin{steps}
	\item On Canvas, navigate to the People section, then to the ``Groups'' tab. Scroll to a group called ``L1 Scale [number]'' that isn't used and have each person in your group add themselves to that same lab group.
\end{steps}

This enables group grading of your lab report. Only one person will submit the group report, and all members of the group will receive the grade and have access to view the graded assignment.

\section{Your current intuition}\label{se:sec:intuition}

In order to gauge your current sense of where things are in relation to each other, each member of your group will first make three different small, qualitative scale models. \textbf{For this section, do not use a book, or the Internet, other people, or any other resource, to guide your efforts.} This will help you see where your current intuition lies. And when you compare to others, do not change your guess --- it is expected that you might not have an accurate conception yet.

\begin{steps}
	\item Individually, without looking at your groupmates' work, take a blank sheet of paper and \textbf{draw} a horizontal line across the page. On the left end of it, place a dot and mark it ``Sun''. On the right end, place a dot and mark it ``Neptune''. Now place and label a dot for your best guess of the orbital distance for each of the seven other planets orbiting the Sun.

	\item Next, \textbf{make a similar scale model drawing}, this time placing the following objects on it: the Sun, Neptune, the Voyager I robotic probe (the furthest human-made object from Earth that started its journey in 1977), our nearest star Proxima Centauri, and its planet, Proxima Centauri b.
	
	\item Finally, \textbf{draw} the Sun and all 8 planets with your best guess of their relative sizes --- you should end up with 9 circles. The distances do not need to be scaled for this estimate.

	\item Compare your drawings with your groupmates. \textbf{Record any surprises or big differences that you had.}
\end{steps}

\section{Making an accurate scale model}\label{se:sec:model}

Now that you have your current sense of it, your lab section will make accurate scale models of the three scales you made above. Bigger is better, so 
you will use the hallway that extends on the second floor from the south end of Kersten, across the skywalk, through Eckhardt, and into the Accelerator Building.
%feel free to get creative in finding a long distance you can work from to make your scale. Only go outside to do so if you are confident you won't increase your risk.

\textbf{Available equipment:} measuring wheel, meter stick, tape measure, paper, masking or label tape, scissors, markers
%\textbf{Possible equipment:} paper, masking or label tape, scissors, markers

%Your group will be assigned one of the 3 scale models described in the previous section to create in the hallway. After you create it, you will make a short ($\sim 5\:$min) video presentation and upload it to a Canvas Discussion, describing what you found and what your impression is, and walking the class through the scale model that you made down the hallway.

\subsection{Tips}
\begin{itemize}
	\item To make a scale model, you will need to gather the size or distance information for the objects in your model using any resource you'd like, then divide each of those by the same number to create your scale. You can use a spreadsheet to make this task easier.
	
	\item For the two distance scale models, you will need to measure the distance you have to work with in the hallway.
	Ensure that you can see from one end to the other.
	Prop open the skywalk doors if needed.
	You should place some kind of upright sign, perhaps taped down, at each spot where an object is in your model.
	
	\item For the size scale model, since all the objects are mostly spherical, you can create your model by cutting out circles of paper. For the larger objects, feel free to tape sheets of paper together to make a bigger circle.
\end{itemize}

\begin{steps}
	
	\item Once all groups are ready to give their tour, take turns visiting each other's scale models and walking through them.
	
	\item Make a table of the real distances and sizes and your scaled down versions. \textbf{Include this in your report.}
	
	\item \textbf{Document your scale model} with pictures and/or video (you can upload your video somewhere and include a link in your lab report).
\end{steps}

\section{Appreciating distances}\label{se:sec:calc}

Answer the following questions:
\begin{enumerate}
	\item Using a nominal speed of a car on Earth, how long would it take to drive:
	\begin{enumerate}
		\item once around the equator?
		\item from the Earth to the Sun?
		\item from the Sun to Neptune?
		\item from the Sun to Proxima Centauri?
	\end{enumerate}
	\item If you wanted to travel each of these distances in 1 year, how fast would you need to go in each case?
	\item For the longest distance, how does that speed compare to the speed of light, which is the fastest anything can go?
\end{enumerate}

\section{Report checklist}

Include the following in your lab report. See Appendix~\ref{cha:lab-report-format} for formatting details.

\begin{enumerate}
	\item Your intuition estimates from Section~\ref{se:sec:intuition} and your reflection of how they compare to your group mates'.
	\item A table of the distances/sizes and scaled version that you created in Section~\ref{se:sec:model}.
	\item Pictures or video of your scale model (Step 9).
	\item Worked solutions to the questions in Section~\ref{se:sec:calc}.
	\item A 100--200 word reflection on the scale --- was there anything that surprised you? How do you see your place in the universe, given the scale of the solar system and how far away even the nearest star is?
	\item Write a 100--200 word paragraph reporting back from each of the four roles: facilitator, scribe, technician, skeptic. Where did you see each function happening during this lab, and where did you see gaps? What successes and challenges in group functioning did you have? What do you want to do differently next time?
\end{enumerate}
