\chapter[Bending light to see into space]{Bending light to see into space\footnote{Initial lab written by Brent Barker. Home measurement setup, photos, and text provided by Chris Varney.}}

%todo feedback from 2019-fall:
% - uncertainty calculation was one step too much for many students. either remove or explicitly make it easy (=STDEV, or just do range of measurements)
% - logistics, mini-lecture, and queue instructions took 30 minutes
% - in step 4, give example table of observations
% - the 2 focal length measurements seemed redundant
% - lab was too long, most students did not get to making the telescope, but those who did thought it was really cool. formula constructing was out of reach for all.
% - students really wanted to know why light was refracting, not just that it was

% - "the wording of the manual is confusing at times. My main confusion was with use of the ambiguous word "object" to describe the ray box. Perhaps a more general term such as 'light source' would be more helpful than the term 'object.' "

Optical telescopes are one way that astronomers use to better observe the cosmos. In this lab, you will build up your understanding about how telescopes help us do this.

\section{Learning Goals}

\begin{itemize}
	\item Learn how light behaves when traveling between mediums.
	
	\item Create an image with a lens.
	
	\item Configure a refracting telescope and explain how it helps for observing the sky.
\end{itemize}

\section{Forming Groups}

Find your group of 2--3 people. You can change groups from week to week, if you'd like.

\begin{steps}
	\item Once you have a group, meet with each other and decide a) what tools you will use to communicate and collaborate, b) when you will meet, c) what you will do when you need to change an agreement, and d) what you will do when a member has a concern about how the group is functioning. \textbf{Record your agreements.}% This part counts as data collection and analysis, so it can be identical in each member's report.}
\end{steps}

\subsection{Team roles}

\begin{steps}
	\item \textbf{Decide on roles} for each group member.
\end{steps}

The available roles are:
\begin{itemize}
	\item Facilitator: ensures time and group focus are efficiently used
	\item Scribe: ensures work is recorded
	\item Technician: oversees apparatus assembly, usage
	\item Skeptic: ensures group is questioning itself
\end{itemize}

These roles can rotate each lab, and you will report at the end of the lab report on how it went for each role. Some members will be holding more than one role. For example, you could have the skeptic double with another role. Consider taking on a role you are less comfortable with, to gain experience and more comfort in that role.

Additionally, if you are finding the lab roles more restrictive than helpful, you can decide to co-hold some or all roles, or think of them more like functions that every team needs to carry out, and then reflect on how the team executed each function.

\subsection{Add members to Canvas lab report assignment group}

\begin{steps}
	\item On Canvas, navigate to the People section, then to the ``Groups'' tab. Scroll to a group called ``L1 Bending Light [number]'' that isn't used and have each person in your group add themselves to that same lab group.
\end{steps}

This enables group grading of your lab report. Only one person will submit the group report, and all members of the group will receive the grade and have access to view the graded assignment.


%\section{But first! Observing with Stone Edge Observatory}
%
%Throughout the quarter, you will be taking data (images) with a robotic telescope that is located in Sonoma, California. This week, you will select a target and queue some images to be taken, at \url{queue.stoneedgeobservatory.com}. Please choose from one of the targets in Table\ \ref{lt:tab:targets}. The camera is a monochrome camera, broadly sensitive to all visible wavelengths ($\sim 400$--$700\:$nm). For now, let's collect light in all of those wavelengths by selecting the ``Clear'' filter, as well as leaving ``Dark'' checked.
%
%\renewcommand{\arraystretch}{1.2}
%
%\begin{table}
%	\begin{tabular}{p{0.6in}p{.8in}p{.9in}p{0.6in}p{0.9in}p{2in}}
%		target & RA & Dec & V-band & Suggested & Notes \\
%		& (hh mm ss) & ($\pm$ dd mm ss) & magnitude & exposure time (s) & \\ \hline
%		51 Pegasi & 22 57 27.9804 & +20 46 07.7822 & 5.46 & 15 & first Main Sequence star found to host an exoplanet \\
%		Altair & 19 50 46.9986 & +08 52 05.9563 & 0.76 & 5 & rapidly rotating star \\
%		
%		Betelgeuse & 05 55 10.3054 & +07 24 25.4304 & 0.42 & 5 & red supergiant star and one of the largest stars visible with the naked eye \\
%
%		Deneb & 20 41 25.91514 & +45 16 49.2197 & 1.25 & 10 & forms the Summer Triangle with Altair and Vega  \\
%	
%		Kelt-9 & 20 31 26.3534 & +39 56 19.7652 & 7.56 & 30 & hosts the hottest known gas giant exoplanet \\
%	
%		Epsilon Eridani & 03 32 55.8450  & -09 27 29.7312 & 3.73 & 15 & one of the nearest Sun-like stars with a planet and very young \\
%	
%		KIC 8462852 (aka Tabby's Star) & 20 06 15.4527 & +44 27 24.7914 & 12.01 & 90 & has very irregular changes in brightness (ask Prof. Fabrycky about what causes the dips!) \\
%	
%		Mira & 02 19 20.7921 & -02 58 39.4956 & 6.53 & 25 & variable binary system with a red giant and a white dwarf \\
%	
%		Polaris & 02 31 49.0946 & +89 15 50.7923 & 2.02 & 10 & lines up almost exactly with Earth's rotational axis to the north and is often called the North Star \\
%	
%		Sirius & 06 45 08.9173 & -16 42 58.0171 & -1.46 & 2 & brightest star in the sky \\
%	
%		UY Scuti & 18 27 36.5286 & -12 27 58.8933 & 11.20 & 90 & largest known star by radius \\
%	
%		Vega & 18 36 56.3364 & +38 47 01.2802 & 0.03 & 5 & one of the most well studied stars in the sky and has been used as a reference for calibrating photometric brightness scale for stars
%
%	\end{tabular}
%	\caption{List of targets for you to choose from.}\label{lt:tab:targets}
%\end{table}
%
%So that the queue is not overwhelmed, please limit your group's observations to one queue submission per student per week.

\section{How do we use materials to bend light?}

We use lenses all the time to shape the path that light takes, either with eyeglasses or with optical telescopes. These activities are intended to help you understand how we use materials to bend light.

\begin{steps}
	\item Find the ray box and a trapezoidal prism in your kit. Set the prism on a white sheet of paper and adjust the ray box so that it is shining 1 ray of line from the side and onto the prism. Play with the angle at which the ray strikes the prism and observe what happens to the light ray that extends into the prism. How does the path of the ray change when it enters the prism? \textbf{Record your observations. Be specific about any patterns you notice.}
	
	\item It's hard to see the light in the prism, so let's use a simulation. Go to \url{https://phet.colorado.edu/en/simulation/bending-light} and select the play button to launch the simulation. Select the leftmost ``Intro'' box. This is a side view of the interface between two materials, currently air on the top half of the screen and water on the bottom. A laser is positioned above the interface.
	
	\item Play with the controls on this screen until you have an idea of how to move the laser, turn it on and off, and adjust the materials on the top and bottom. To reset the simulation, select the orange circular button on the lower right.
	
	\item Observe what happens to the laser beam when you change the angle, materials, and indices of refraction. \textbf{Record your observations for your lab report.} For example, ``when the angle between the laser and the interface gets smaller, the laser gets deflected more/less''.
	
	\item Open the second screen at the bottom of the sim. Lenses are wider in the middle and thinner at the top and bottom, so drag a triangle up and shine a ray through it. Then bring a second triangle up, rotate it upside down, and arrange the two so they look a bit like a convex lens. Set the laser to output several parallel rays and aim it so some of the rays hit the top triangle and some hit the bottom triangle. \textbf{Record your observations.}
	
	\item Go back to your physical ray box and optics set. Adjust the ray box to emit 5 parallel rays and set up the convex lens (the one that is thinner on the ends and thicker in the middle) so that the rays are hitting it from the side. Notice where the rays go after they go through the lens. \textbf{Record your observations.}
\end{steps}

This setup of parallel rays is convenient for seeing precisely what the lens does. It also happens to be the situation when we observe things that are very far away compared to length scales of the lens. Consider two stakes driven into the ground next to each other, both perpendicular to the ground (and thus pointing directly at the center of the Earth). Since the Earth is a sphere, those stakes can't actually be both pointing directly toward the center of the Earth and also parallel to each other. For the former to be true, they must be angled slightly away from each other. But since the distance between them is so short compared to the distance away from the Earth's center, they are effectively parallel. \textit{This is the same with light arriving from distant objects like stars.}

\begin{steps}
	\item Let's try putting some parallel rays from a distant object on a lens and see what happens where those rays intersect. Take one of the round lenses in a black plastic lens holder and shine those parallel rays from the light box on it to find the distance where the rays intersect.
	
	\item Select a distant bright object with sharply contrasting edges, like a building outside the window during the daytime, an exposed florescent bulb, or the light-up target on the other side of the ray box. Hold the lens between the object and a white sheet of paper. The white sheet of paper should be placed about the same distance away as the intersection distance. \textbf{Record your observations.}
\end{steps}

This intersection distance is called the ``focal length'' of the lens, and if an object is a long distance away compared to the focal length, then its image is formed at the focal length (if the object is closer, the the image is formed further away than the focal length).

This principle works in reverse too --- if an object is placed at the focal length of the lens, the rays come out parallel on the other side. The image is effectively formed an infinite distance away.

\begin{steps}
	\item Design and conduct an experiment to find the focal length of the lens you just used. Decide as a group how to measure, and how to estimate an uncertainty for your measurement. See Appendix\ \ref{cha:uncertainty} for detailed information about estimating uncertainty. \textbf{Record a sketch of your setup, a description of your procedure for gathering and analyzing your data, and the data itself.}
	
	\item Compare: how does this focal length compare to the focal length that is printed on the lens holder? See Appendix\ \ref{unc:sec:comparing} for how to compare two values, taking into account their uncertainties. \textbf{Record this comparison calculation and what you conclude about how close they are. Is the printed value correct?}
\end{steps}

This lens setup is great for producing images, for example to record onto photographic film or a digital camera's image sensor. It's less good for look through, to magnify and gather more light the way we want to with a telescope. For that, we'll need at least two lenses.

\section{Your first telescope}

Telescopes come in many different configurations. Here you'll construct one that is simple by comparison, a refracting telescope, using just 2 lenses.

\begin{steps}
	\item Here's the principle for building this telescope: the image created by the first lens, called the objective lens, is the object for the second lens, which is called the eyepiece lens. The thing we are wanting to look at (the object for the objective lens) is far away. We want the image created by the eyepiece lens to be an infinite distance away on the near side (just trust me on this). \textbf{Given these design goals and the information about lenses above, where should the two lenses be positioned with respect to each other? Sketch your proposal, labeling each lens and drawing the focal lengths of each lens.}
	
	\item Using the optical bench, construct and test your telescope. Look at a distant (across the room) object through it. If you don't get a clean image of it by looking through the eyepiece lens, iterate on your design until you get it.
	
	\item Find the magnification of your telescope. Hint: if you have two identical objects, how close does one need to be to look the same size/distance as the one you see through the telescope? Experimentally determine this magnification factor (1 is no magnification, 2 means it looks twice as close, 0.5 means that it looks twice as far away, etc), and estimate the uncertainty of your magnification. \textbf{Record this.}
	
	\item Magnification should be related to the properties of the two lenses. Make up a formula that relates the focal lengths of your lenses to the magnification of your telescope. You may need to switch the lenses around or use different ones to test your formula. \textbf{Record this.}
\end{steps}

\section{Report checklist and grading}

Each item below is worth 10 points. See Appendix\ \ref{cha:lab-report-format} for guidance on writing the report and formatting tables and graphs.

\begin{enumerate}
	
	\item Detailed observations from Steps 4--11.
	
	\item Sketch of your setup and procedure for finding the focal length in Step 12, as well as value, with uncertainty, of the focal length.
	
	\item Comparison of your measured focal length to the manufacturer's stated focal length, from Step 13.
	
	\item Detailed sketch of your working telescope design, from Steps 14--15.
	
	\item Experimental determination of the magnification of your telescope, with uncertainty (Step 16).
	
	\item Formula relating the focal lengths to the magnification (Step 17).
	
	\item Discuss the findings and reflect deeply on the quality and importance of the findings. This can be both in the frame of a scientist conducting the experiment (``What did the experiment tell us about the world?'') and in the frame of a student (``What skills or mindsets did I learn?'').
	
	\item Write a 100--200 word paragraph reporting back from each of the four roles: facilitator, scribe, technician, skeptic. Where did you see each function happening during this lab, and where did you see gaps? What successes and challenges in group functioning did you have? What do you want to do differently next time?
\end{enumerate}


%\begin{itemize}
%	\item Detailed observations from Steps 1--8.
%	
%	\item Sketch of your setup and procedure for finding the focal length in Step 9, as well as value, with uncertainty, of the focal length.
%	
%	\item Comparison of your measured focal length to the manufacturer's stated focal length, from Step 10.
%	
%	\item Detailed sketch of your working telescope design, from Steps 11--12.
%	
%	\item Experimental determination of the magnification of your telescope, with uncertainty.
%	
%	\item Formula relating the focal lengths to the magnification.
%	
%	\item Discuss the findings and reflect deeply on the quality and importance of the findings. This can be both in the frame of a scientist conducting the experiment (``What did the experiment tell us about the world?'') and in the frame of a student (``What skills or mindsets did I learn?'').
%	
%\end{itemize}