\begin{landscape}
\chapter{Rubrics}\label{cha:rubrics}

\begin{longtable}{>{\bfseries}p{0.04\textheight}|>{\bfseries\RaggedRight}p{0.23\textheight}|>{\RaggedRight}p{0.21\textheight}|>{\RaggedRight}p{0.21\textheight}|>{\RaggedRight}p{0.22\textheight}|>{\RaggedRight}p{0.22\textheight}}
	\toprule
	& Scientific Ability
	& Missing & Inadequate & Needs Improvement & Adequate \\ \midrule \endhead
	A11
	& Graph
	& No graph is present.
	& A graph is present but the axes are not labeled. There is no scale on the axes.
	& The graph is present and axes are correctly labeled, but the axes do not correspond to the independent and dependent variables, or the scale is not accurate.
	& The graph has correctly labeled axes, independent variable is along the horizontal axis and the scale is accurate.
	\\
	\bottomrule
	\caption{Rubric A: Ability to represent information in multiple ways}
	\end{longtable}

\begin{longtable}{>{\bfseries}p{0.02\textheight}|>{\bfseries\RaggedRight}p{0.25\textheight}|>{\RaggedRight}p{0.21\textheight}|>{\RaggedRight}p{0.21\textheight}|>{\RaggedRight}p{0.22\textheight}|>{\RaggedRight}p{0.22\textheight}}
		\toprule
		& Scientific Ability
		& Missing & Inadequate & Needs Improvement & Adequate \\ \midrule \endhead
		B1
		& Is able to identify the phenomenon to be investigated
		& No phenomenon is mentioned
		& The description of the phenomenon to be investigated is confusing, or it is not the phenomenon of interest.
		& \midsloppy The description of the phenomenon is vague or incomplete.
		& The phenomenon to be investigated is clearly stated. \\ \midrule
		B2
		& Is able to design a reliable experiment that investigates the phenomenon
		& The experiment does not investigate the phenomenon.
		& The experiment may not yield any interesting patterns.
		& Some important aspects of the phenomenon will not be observable.
		& The experiment might yield interesting patterns relevant to the investigation of the phenomenon. \\ \midrule
		B3
		& Is able to decide what physical quantities are to be measured and identify independent and dependent variables
		& The physical quantities are irrelevant.
		& Only some of physical quantities are relevant.
		& The physical quantities are relevant. However, independent and dependent variables are not identified.
		& The physical quantities are relevant and independent and dependent variables are identified. \\ \midrule
		B4
		& Is able to describe how to use available equipment to make measurements
		& At least one of the chosen measurements cannot be made with the available equipment.
		& All chosen measurements can be made, but no details are given about how it is done.
		& All chosen measurements can be made, but the details of how it is done are vague or incomplete.
		& All chosen measurements can be made and all details of how it is done are clearly provided. \\ \midrule
		B5
		& Is able to describe what is observed without trying to explain, both in words and by means of a picture of the experimental setup.
		& No description is mentioned.
		& A description is incomplete. No labeled sketch is present. Or, observations are adjusted to fit expectations.
		& A description is complete, but mixed up with explanations or pattern. Or the sketch is present but is difficult to understand.
		& Clearly describes what happens in the experiments both verbally and with a sketch. Provides other representations when necessary (tables and graphs). \\ \midrule
		B6
		& Is able to identify the shortcomings in an experiment and suggest improvements
		& No attempt is made to identify any shortcomings of the experiment.
		& The shortcomings are described vaguely and no suggestions for improvement are made.
		& Not all aspects of the design are considered in terms of shortcomings or improvements.
		& All major shortcomings of the experiment are identified and reasonable suggestions for improvement are made. \\ \midrule
		B7
		& Is able to identify a pattern in the data
		& No attempt is made to search for a pattern.
		& The pattern described is irrelevant or inconsistent with the data.
		& The pattern has minor errors or omissions. Terms like ``proportional'' used without clarity, e.g.\ is the proportionality linear, quadratic, etc.
		& The pattern represents the relevant trend in the data. When possible, the trend is described in words. \\ \midrule
		B8
		& Is able to represent a pattern mathematically (if applicable)
		& No attempt is made to represent a pattern mathematically.
		& The mathematical expression does not represent the trend.
		& No analysis of how well the expression agrees with the data is included, or some features of the pattern are missing.
		& The expression represents the trend completely and an analysis of how well it agrees with the data is included. \\ \midrule
		B9
		& Is able to devise an explanation for an observed pattern
		& No attempt is made to explain the observed pattern.
		& An explanation is vague, not testable, or contradicts the pattern.
		& An explanation contradicts previous knowledge or the reasoning is flawed.
		& A reasonable explanation is made. It is testable and it explains the observed pattern. \\
		\bottomrule
		\caption{Rubric B: Ability to design and conduct an observational experiment \cite{etkina_scientific_2006}.}\label{rubric:b}
	\end{longtable}


\begin{longtable}{>{\bfseries}p{0.02\textheight}|>{\bfseries\RaggedRight}p{0.25\textheight}|>{\RaggedRight}p{0.21\textheight}|>{\RaggedRight}p{0.21\textheight}|>{\RaggedRight}p{0.22\textheight}|>{\RaggedRight}p{0.22\textheight}}
	\toprule
	& Scientific Ability
	& Missing & Inadequate & Needs Improvement & Adequate \\ \midrule \endhead	
	F1
	& Is able to communicate the details of an experimental procedure clearly and completely
	& Diagrams are missing and/or experimental procedure is missing or extremely vague.
	& Diagrams are present but unclear and/or experimental procedure is present but important details are missing. It takes a lot of effort to comprehend.
	& Diagrams and/or experimental procedure are present and clearly labeled but with minor omissions or vague details. The procedure takes some effort to comprehend.
	& Diagrams and/or experimental procedure are clear and complete. It takes no effort to comprehend.
	\\ \midrule
	F2
	& Is able to communicate the point of the experiment clearly and completely
	& No discussion of the point of the experiment is present.
	& The experiment and findings are discussed but vaguely. There is no reflection on the quality and importance of the findings.
	& The experiment and findings are communicated but the reflection on their importance and quality is not present.
	& The experiment and findings are discussed clearly. There is deep reflection on the quality and importance of the findings.
	\\
	\bottomrule
	\caption{Rubric F: Ability to communicate scientific ideas \cite{etkina_scientific_2006}.}\label{rubric:f}
\end{longtable}

\begin{longtable}{>{\bfseries}p{0.02\textheight}|>{\bfseries\RaggedRight}p{0.25\textheight}|>{\RaggedRight}p{0.21\textheight}|>{\RaggedRight}p{0.21\textheight}|>{\RaggedRight}p{0.22\textheight}|>{\RaggedRight}p{0.22\textheight}}
	\toprule
	& Scientific Ability
	& Missing & Inadequate & Needs Improvement & Adequate \\ \midrule \endhead	
	G1
	& Is able to identify sources of experimental uncertainty
	& No attempt is made to identify experimental uncertainties.
	& An attempt is made to identify experimental uncertainties, but most are missing, described vaguely, or incorrect.
	& Most experimental uncertainties are correctly identified. But there is no distinction between random and instrumental uncertainty.
	& All experimental uncertainties are correctly identified. There is a distinction between instrumental and random uncertainty.
	\\ \midrule
	G2
	& Is able to evaluate specifically how identified experimental uncertainties affect the data
	& No attempt is made to evaluate experimental uncertainties.
	& An attempt is made to evaluate uncertainties, but most are missing, described vaguely, or incorrect. Or the final result does not take uncertainty into account.
	& The final result does take the identified uncertainties into account but is not correctly evaluated. Uncertainty propagation is not used or is used incorrectly.
	& The experimental uncertainty of the final result is correctly evaluated. Uncertainty propagation is used appropriately.
	\\ \midrule
	G3
	& Is able to describe how to minimize experimental uncertainty and actually do it
	& No attempt is made to describe how to minimize experimental uncertainty and no attempt to minimize is present.
	& A description of how to minimize experimental uncertainty is present, but there is no attempt to actually minimize it.
	& An attempt is made to minimize the uncertainty in the final result is made but the method is not very effective.
	& The uncertainty is minimized in an effective way.
	\\ \midrule
	G4
	& Is able to record and represent data in a meaningful way
	& Data are either absent or incomprehensible.
	& Some important data are absent or incomprehensible. They are not organized in tables or the tables are not labeled properly.
	& All important data are present, but recorded in a way that requires some effort to comprehend. The tables are labeled but labels are confusing.
	& All important data are present, organized, and recorded clearly. The tables are labeled and placed in a logical order.
	\\ \midrule
	G5
	& Is able to analyze data appropriately
	& No attempt is made to analyze the data.
	& An attempt is made to analyze the data, but it is either seriously flawed or inappropriate.
	& The analysis is appropriate but it contains errors or omissions.
	& The analysis is appropriate, complete, and correct.
	\\
	\bottomrule
	\caption{Rubric G: Ability to collect and analyze experimental data \cite{etkina_scientific_2006}.}\label{rubric:g}
\end{longtable}

\end{landscape}